# (PART) Introduction {-} 

# Data Science

Data science is a field of science that aims at discovering patterns and gaining knowledge through analysis of data and developing methods for support such investigation.
The field has its foundation in mathematics, statistics, computer science, and vision science and contributed to a wide array of disciplines including biology chemistry, economics, literary studies, medicine, physical sciences, psychology, etc.

What are examples of data science?
Here is a short list (each item with a possible academic discipline that appears the most relevant to it):

* Tracking places where crimes occur in cities for crime prevention (criminology)
* Tracking social media postings to understand changes in public opinions (political science)
* Dynamically changing advertisements in search engines based on the browsing history and query terms (marketing)
* Predicting trajectories of hurricanes (physical sciences)
* Finding genetic factors contributing diseases (medicine)
* Differentiating among novelists based on their writing styles (literary studies)
* Winnowing an enormous candidate list of chemicals for a target effect based on their known properties (chemistry)
* Diagnosing pulmonary diseases based on patient X ray images (medicine)
* Identifying students that are at-risk (education)

The prominence of data science comes from the fact that the technologies for collecting, sharing and analyzing data have become widely accessible.
In many disciplines, research by way of data science has received a formidable status.
Data science has gained a formidable status is in many disciplines.
It is thus important for new generations of professionals (educators, scientists, engineers, doctors, etc.) to know how data sciences and know how to start data science explorations.
The purpose of this course is to introduce its students to the very basic of data science.

We will use the programming language R in our exploration.
R is a programming language with roots in statistics and has a wide variety to tools that make it easy for its user to conduct statistical analysis and visualization.
The syntax for R (i.e., the rules for writing the code for R) is more flexible than many other programming languages.
It also allows its use to create a "notebook", interleaving non-programming parts and programming parts in an intuitive manner.
Also, R can sit in the computer of its user.
Without depending on external sources (except when loading programs other people developed for future use and loading data sets from internet sources).
Thus, after completing the course, the students will be able to keep running and modifying their courses with new data sources and new ideas.

### A Tiny Example

In this course, we use a package `tidyverse`.
A package is a group of source codes that serve a certain purpose.
A nonprofit organization promoting R, [CRAN](https://cran.r-project.org/), maintains packages researchers have developed as a resource base for providing them.
The package `tidyverse` is a collection of packages, which we call a "meta-package", that is useful for data analysis and data visualization.

Here is an example of plotting data using `tidyverse`.
R comes with a number of data sets.
One of them is `WordPhones`, which provides the number of telephones in some large geographical regions of the world in 1951 and 1956-1961.

The analysis begins by loading the package into your programming environment.

```{r}
library(tidyverse)
```

`tidiverse` uses a specific format for data bases.
The data set is native to R, not to `tidyverse`, and so we need to load the data set into a format that is processible with `tidyverse`.


```{r}
wp <- as.data.frame(WorldPhones)
wp
```

Let us say our goal is to plot the trend in the seven regions over the years.
To accomplish the goal, we need to transform the data so that the area and the number appear as column pairs.
Also, we change the row names "1951", etc., to the values of a column.

```{r}
wp0 <- wp %>% rownames_to_column("Year") %>% pivot_longer(cols = 2:8, names_to = "Area", values_to = "Count")
# wp1 <- as.data.frame(t(wp)) %>% rownames_to_column("Area")
wp0
```

We can now plot the changes in the numbers with the year as the horizontal axis and the count as the vertical axis.
We assign different colors to the seven areas using the automatic coloring functionality of the plotting function.
We see that the numbers steadily increased in North America, Europe, and Asia.

```{r}
ggplot(wp0, aes(x = Year, y = Count)) +
  geom_point(aes(color = Area))
```

We can now plot the chnages as bars where the same colors represent the regions.

```{r}
ggplot(wp0) +
  geom_bar(aes(x = Year, y = Count, fill = as.factor(Area)), stat = "identity") +
  labs(fill = "Area")

```

### Text Processing


You can download from a web site.
[The Gutenberg Project](http://www.gutenberg.org/) is an international project to provide literary texts in a variety of formats for the public use.
One of the main literary works the site offers is "Moby Dick" by Herman Melville.

```{r}
moby_url <- "http://www.gutenberg.org/files/2701/2701-0.txt"
moby_text <- readtext(moby_url)
main <- moby_text[1,2]
```


The first 1,000 characters of the text are as follows:

```{r}
str_sub(main, 1, 1000)
```

Notice that the text has a header where "\n" represents the return key.
The real text of the work starts with "  ETYMOLOGY." and ends before "End of Project Gutenberg".
We extract every character between the the positions these two signature strings spedcify.

```{r}
p_start <- str_locate(main, fixed("  ETYMOLOGY."))[1]
p_start
p_last <- str_locate(main, fixed("End of Project Gutenberg"))[1]
p_last
main0 <- str_sub(main, p_start, p_last - 1)
```

The main text starts with "\\nCHAPTER ".
We use it as the separator and generate a list of tests, where each element of the list is a chapter.
The `length(main_parts)` shows the number of elements in the list.

```{r}
main_parts <- unlist(str_split(main0, "\\nCHAPTER "))
length(main_parts)
```

We turn the list of chapter texts into a data set.
We convert it to all lowercase and replace the return key to the white space.
Let us see the first 1,000 characters of the 100th chapter.
Since the work starts with an epilog, the actual chapter number is off by 1, i.e., 99.
The first three characters, "99.", represent the chapter number.



```{r}
main_data <- as.data.frame(main_parts)
main_data %>% rename(original = main_parts) %>% transmute(origLower = str_replace_all(str_to_lower(original), "\\n", " ")) -> main_data0
str_sub(main_data0$origLower[100],1,1000)
```

```{r}
colnames(main_data0)
main_data0 %>% mutate(len = str_length(origLower)) -> main_data1
colnames(main_data1)
#main_data0$len
```

We assigned the name "origLower" to the all-lowercase text we generated.
We compute in each chapter how many times the following words occur: "whale", "ship", "sea", "this", "that", and "god".

```{r}
main_data1 %>% mutate(whale = str_count(origLower, "whale")) %>% mutate(ship = str_count(origLower, "ship"))%>% mutate(sea = str_count(origLower, "sea")) %>% mutate(this = str_count(origLower, "this")) %>% mutate(that = str_count(origLower, "that"))%>% mutate(god = str_count(origLower, "god")) -> main_data2
```

We now plot the relationships between "this" and "that" and the relationships between "whale" and "god".
In the first, we see that where "this" appears a lot "that" appears a lot as well.
In the second, we see that the occurrences of "god" have nothing to do with the occurrences of "whale", except there is one chapter with a very high number of occurrences of "god".

```{r}
ggplot(main_data2, aes(x = this, y = that)) +
  geom_point() + geom_smooth()
ggplot(main_data2, aes(x = whale, y = god)) +
  geom_point() + geom_smooth()
```


### A Rough Organization

Here is the plan for the course.
We will first learn the basics of R programming.
We will then learn how to use `tidyverse` for data formatting and data plotting.
After that, we will study a variety of techniques for analyzing data and visualizing the results.

We hope you will find our journey interesting.






<!--
```{r}
library(tidyverse)
library(readtext)
```

```{r}
scarlet_url <- "https://dev.gutenberg.org/files/36/36-0.txt"
scarlet_text <- readtext(scarlet_url)
head(scarlet_text)
```


```{r}
main <- scarlet_text[1,2]
str_sub(main, 100, 200)
```

```{r}
str_locate(main, fixed("BOOK ONE\nTHE COMING OF"))
str_locate(main, "\\n[IXVL][IXVL]*\\.\\n")
```

```{r}
main_parts <- unlist(str_split(main, "\\n[I|X|V|L][I|X|V|L]*\\.\\n"))
length(main_parts)
for (i in 1:28) {
  print(str_sub(main_parts[i], 1, 20))
}
lll <- str_length(main_parts)
lll
```

```{r}
main_data <- data.frame(as.matrix(lll)) # %>% rename("original" = 1) %>% mutate(first = str_sub(original, 1, 100), .keep = "unused")
```

```{r}
nrow(main_data)
```
-->

