---
output:
  pdf_document: default
  html_document: default
---
# (PART) Programming Fundamentals {-} 

# Introduction 

The work of a data scientist consists of many aspects:

* working with domain experts to determine what questions need addressing
* assembling a group of data into a dataset for analysis
* examining the dataset to remove any unusable parts
* analyzing the data to addressing the research question(s)
* sharing the results with domain experts

Some of these steps are quite mechanical or cumbersome, and may repeat the same calculation again and again. Mistakes are costly; they can lead to wrong conclusions and errors on the job can mean money lost. It would be best not to butcher your data science job because of simple mistakes.

However, to err is human. So how can you minimize the chance of those errors occurring? That's where computer programming come to the rescue.

## Computers

A computer program is a sequence of characters with which to tell a computer how to perform its task. The history of computers is kind of long. Early in the 1800s, Charles Babbage invented a machine for calculation, calling it the ["Difference Engine"](https://www.computerhistory.org/babbage/engines/). Ada Lovelace, a daughter of English poet Lord Byron, collaborated with Babbage to develop ideas about what the engine could compute beyond addition and multiplication.

Several attempts to build such mechanical systems to perform calculation had emerged until the first half of the 1900s during which a series of innovations resulted in the invention of "computers". The new inventions used electricity as the source of energy, vacuum tubes for regulating electrical currents, separation between the "brain" that is responsible for computation and the storage for holding the final and intermediate results of computation, and punch cards for defining the calculation steps to follow.

### Programming languages and Scripting Languages

In the early days of modern computing, programmers had to provide instructions to computers using languages specific to the computers. However, as new computers emerged and the number of computers in use grew, there arose a need for languages that programmers can use in describing the instructions.

That is the birth of modern, machine-independent programming languages. Allowing the use of machine-independent languages in writing programs requires the creation of a layer that sits between programmers who code with universal languages and the computers who will execute the program but do not understand the languages the programmers use.

Early on, there was only one layer separating programmers from computers. Nowadays there are multiple layers separating between them. What these layers do is outside the coverage of the course, and we will skip it entirely. The key point is that a programmer can use a specific language to write what she intends to do with program.

Many types of programming languages exist. One type is the *scripting language*. In a scripting language, a programmer "enters" an environment specific to the language and then executes her program "line by line". 

A *line* of a program is a sequence of characters which is syntactically complete according to the syntax of the language. In scripting language programming, a programmer completes writing of a line by pressing the *return key* (and the like), and then the environment immediately parses the line the programmer has entered regarding the syntax of the language and performs any action the line expresses.

### Programming in R

We will use the programming language [R](https://www.r-project.org/) in this course.
There are four major reasons to choose R as our language.

* R is quite popular. There are many data scientists out there who enjoy conducting science using R.
* There are so many R codes for public use. You can adapt such programs to your own needs. Reciprocally, you can share your R codes with others. Some of the popular codes have become libraries, groups of R codes that you can use for specific purposes. You can download such libraries for your own use with a simple line of code.
* R comes with dozens of datasets that you play with.
* There exists an easy-to-install R programming environment [RStudio](https://rstudio.com/products/rstudio/download/), where not only can you write code but also can create notebooks in which R code and prose alternate.

## Expressions

In R, executable "lines" must conform to the syntax of the language. If a programmer enters a syntactically incorrect line, the environment usually issues an "error message" with some information about why the line failed.

Entering an incorrect line has no effect on the programming task she wants to carry out, which means there is nothing to be afraid of about making mistakes!

A line of an R code consists of what we call *expressions*. Several types of expressions exist, which serve as building blocks. We will introduce the types as they become necessary.

Let us begin with the simple ones: numerals and arithmetic expressions.

### Numerical Literals

*Numerical literals* are expressions of numbers with exact specifications of their values. There is no ambiguity in the expressions.

Here are some examples of numerical literals. 

```{r}
415
-3.56
1956.5436781
```

An important numerical literal type uses `e` to specify shifting the decimal point. You are probably already familiar with this idea from other courses. It is called writing in *scientific notation*. 

```{r}
5.07e-2
```

which is equivalent to $5.07 * 10^{-2}$. 

A slight detail that you may want to keep in mind is that each value carries a fixed number of digits that it can correctly represent, and so you cannot be too specific about the value. Consider the following tiny value. 

```{r}
1.5454786e-10
```

The value turns out to be slightly off! 

### Arithmetic Expressions

You can construct mathematical expressions combining numerical literals and mathematical operators `+`, `-`, `*`, and `/` as we use in mathematics. Also, `^` represents exponentiation.

The same order of operations apply when processing an expression with multiple types of operations in it. You can use parentheses, `(` and `)`, to change the order. We often use other types of parentheses like `[` and `]` in writing mathematical expressions in English. However, these carry different meaning in R and so cannot be used in writing mathematical expressions.

When you type an arithmetic expression in an R environment and press the return key, R attempts to evaluate the expression. For example:

```{r}
3.4 * (1.7 + 2.3)
```

```{r}
2^5
```

Instead of using `^` to mean exponentiation, you can also use `**`. Note that there should not be a space between the two `*`.

```{r}
2 ** 5
```

Another important operation is `%%`, which is the *remainder*.

```{r}
10.0 %% 2.3
```

After subtracting 2.3 from 10.0 four times, we arrive at 0.8, which is the remainder.


<!--Let X and Y be two doubles such that Y is not equal to 0.
The value of the remainder of X divided by Y is the remainder when we keep either subtracting from X or adding to X the value of Y until the absolute value we generate from X is smaller than the absolute value of Y. -->

Below is a summary of the mathematical operations.

| Expression    | Operator    |  Example   |
| --------------| ----------- |----------- | 
| Addition      | `+`              |  `10 + 1`  | 
| Subtraction   | `-`              |  `10 - 1`  |
| Division      | `/`              |  `10 / 1`  |
| Multiplication   | `*`           |  `10 * 1`  |
| Remainder   | `%%`               |  `2 - 3`   |
| Exponentiation   | `^`  or `**`  |  `2^3` or `2**3`  |


Here is an example that shows the difference of having parentheses in the expression. Try it by hand first before seeing the answer!

```{r}
2 + 3 * 4 * 5 / 6 ** 2
```

```{r}
2 + (3 * 4 * 5 / 6) ** 2
```

### Giving Names to Results

When you enter a mathematical expression like the one we presented in the above, R responds by printing the value resulting from its evaluation of the expression, but you cannot recall the value. Meaning, we can easily compute the value of the following expression. 

```{r}
3.4 + 4.2
```

But we have no way of *storing* that result! The problem is overcome by *assigning* a name to the value R produces so that we can keep it for future use. 

```{r}
a <- 3.4 + 4.2
```

We can check its value by typing its name. 

```{r}
a
```

Once you have executed it, the assignment will persist until you make another assignment to `a` or you terminate the R environment program you are running. 

```{r}
a <- 5 * 15
a
```

Note how the value of `a` has changed from `7.6` to `75`. We call such named entities that hold a value `objects`.

> You may have noticed the leading `[1]` in front of the value of `a`. This involves the internal workings of R and is a technical detail, so we ignore it for now. 

What happens when we try to refer to an object that currently does not exist? 

```{r, error=TRUE}
barnie
```

Our first programming error! The message reported is quite informative: it states that in the present session there is no assigned value to the name `barnie`.

There are some rules about what kinds of names you can give to an object. A name can use any letter from the alphabet, both uppercase and lowercase, numerals, and the underscore `_`. Underscores are useful when you want to include more than one word in a name. 

```{r}
a_tasty_pie <- 314159
a_tasty_pie
```

A name may not contain spaces or start with a numeral.

### Using Objects in Expressions 

When an R environment evaluates an expression and encounters a name, it tries to see whether an object by the name exists and, if so, substitutes all of its occurrences in the expression with the present value of the object. 

```{r}
a_tasty_pie / 2
```

If the object cannot be found, the environment produces an error message.

```{r error=TRUE}
a <- barnie * 3
```


### Function Calls

Another type of expression is a *function*, which has the same meaning as the *functions* you learned in math class: something that takes some input, does something with it, and produces an output. In R, a function is an object which takes a certain number of objects/values for its execution. 

Here is an example. This function returns the absolute value of a number. 

```{r}
abs(-20)
```

Here are some more examples. 

```{r}
a <- -10.5
floor(a)
ceiling(a)
abs(a)
```

<!--The expression to call a function for its action is `NAME(ARGS)`, where `NAME` is the name of the function object and `ARGS` is the series of object that the function uses in its calculation.
The `ARGS` part appears as a list of objects a comma in between. We call them *arguments*. Arguments are order dependent. --> 

R ships with many functions that we are familiar with from math courses, e.g., `sin`, `cos`, `tan`, `asin`, `acos`, `atan`, `log`, and `log10`, which are sine, cosine, tangent, inverse sine, inverse cosine, inverse tangent, natural logarithm, and logarithm base 10, respectively. The unit of the argument is radian for the first three (i.e., 360 degrees are equal to $\pi$ radians).

```{r}
sin(1)
cos(1)
tan(1)
asin(-1)
acos(-1)
atan(1)
log(10)
log10(10)
```

Functions can of course take more than one argument. The `max` and `min` functions are those that return the maximum and the minimum of the argument list, respectively.

```{r}
max(4, 5, 3)
min(1, 2, 3, 4, 5, 6)
```

These functions accept any number of arguments.

### Example: Pay Your Debt

You may have heard from friends and family the importance of paying debt. You can use R to do a hypothetical calculation of how quickly your debt would accumulate if you do not pay it.

Suppose you load \$1,000 from an agency with a monthly interest rate of 1%, which means that after each month, not only do you still owe the money you owed one month ago but also you owe one percent of the money you owed. 

For example, after the first month of the mortgage in question, you not only owe the \$1,000 you loaned but also 1 percent of it, \$10. Altogether, at the end of the first month, you owe a total of \$1,010.

If you fail to repay, the same series of actions will happen after another month, increasing your total debt to \$1,020.10 dollars. If you fail to repay again, the same series of actions will happen.

Let us see how this process balloons your account if you __DO NOT__ pay for one year.

```{r}
1000 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01
```

The loan accumulate a little over 126 dollars in interest.

The amount does not appear to be much. However, if the "principal amount* is large, the effect is substantial. For example, consider what happens in a loan of \$200,000.

```{r}
200000 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01 * 1.01
```

That's \$25,365 dollars in interest! 

The expression we used had 12 repetitions of 1.01. Using exponentiation, we can simplify it.

```{r}
200000 * 1.01 ^ 12
200000 * 1.01 ** 12
```

Usually when you borrow money from a bank or a mortgage company, you set up a plan for repaying so that you pay the same amount throughout your mortgage. In other words, you pay a fixed amount at every anniversary (usually every month since the start). The bank or the mortgage company subtracts your payment from the amount of money you owe, and one month later will apply the interest rate to the money you now owe. The amount you pay monthly must be so that you owe no more or no less when you make the last payment.

There is a formula that gives the monthly payment.

$$
\mathrm{monthly} = 
\frac{\mathrm{principal} * \mathrm{rate}^{\mathrm{months}} * (\mathrm{rate} - 1)}{\mathrm{rate}^{\mathrm{months}}-1}
$$

Here $\mathrm{principal}$ is the money you borrow, $\mathrm{rate}$ is the monthly rate, and $\mathrm{months}$ is the duration of the loan in terms of the number of months. 

Let us make R do the computation for us.

```{r}
principal <- 1000
rate <- 1.01
months <- 12
monthly <- principal * rate ^ months * (rate - 1) / (rate ^ months - 1)
total <- monthly * months
```

The objects `principal`, `rate`, `months`, and `monthly` represent the four pieces of information appearing in the equation and `total` represents the total payment, which is simply `monthly * months`.

What are the values of `monthly` and `total`?

```{r}
monthly
total
```

We talked about the possibility of not paying for one year. Let us use an object `balloon` to represent the amount.

```{r}
balloon <- principal * rate ^ months
balloon
```

Assuming that in the case of not paying for one year, you pay all the debt, what is the difference between the two options, and what is the proportion of the difference to the principal? Let us use objects `diff` and `diff_percent` to represent the quantities.

```{r}
diff <- balloon - total
diff_percent <- diff / principal * 100
diff_percent
```

So it is a little over 6% that you will have to pay more. 

In other words, we obtain the conclusion:

* Assuming the rate is 1% per month and the loan is for 12 months, if you chose to pay out after 12 months paying none until then, you would have to pay slightly more than 6% of the principal than you would pay with the standard monthly payment program.   

6% does not sound like a significant amount if the principal is small, but what if the principal is 20,000 dollars? Try it out!

## Packages in R

The designers of R had statistical computing in mind. Because of the motivation, many mathematical functions are available as we have seen in the above. There are also, as mentioned earlier, groups of codes that you can use when necessary. We call these *packages*. Here is a list of [R packages](https://cran.r-project.org/web/packages/available_packages_by_name.html); the list is quite long.

Using an R package comes in two stages.

In the first stage, you download the package on to your computer where the R environment can see. When you download an R environment and install it, it sets up the place for it to hold the packages, so there is no need to worry about remembering where they are. The process is by calling the function `install_packages(...)` where the part `...` represents its argument, which is the name of the package. 

In the second stage, the installed packages are now readily usable. When you use one, you have tell the R environment you are using that package. The process is by calling the function `library(...)` where the part `...` represents its argument, which is the name of the package.

Here are the two steps in a nutshell: to install a package, you enter `install.packages(PACKAGE_NAME)`; to use it, you enter
`library(PACKAGE_NAME)`.

A package is a collection of objects. Each object has a name. Sometimes two packages contain objects having the same names. If such objects appear, the R environment warns you about the duplicate names, and chooses to use the most recent name assignment ignoring the earlier ones. Also, if you attempt to load a package with the `library` function and if the package needs installation, the R environment gives you a warning.
Once you have installed the package, the warning will go away. You can also install packages as many times you want.

The package we will most frequently use in this course is [tidyverse](https://www.tidyverse.org/), which is actually a suite of packages designed for doing data science. Once you have installed it, loading the package is as we stated earlier.

```{r, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

### What is the `tidyverse`?

The tidyverse is a collection of packages, each having to do with data visualization and processing. The packages in the suite include:

* `stringr` for manipulating string attributes
* `readr` for reading data from data files
* `tibble` for creating tables and making simple modifications
* `dplyr` for data manipulation (e.g., filtering, reordering, etc.)
* `tidyr` for tidying up data tables
* `ggplot2` for visualization (e.g., creating graphs)
* `purrr` for data mapping and processing nested data

The list seems quite long, but don't worry; we will introduce them gradually throughout the course and only as needed. Each package should be thought of as a "lego block" that works in unison with the other packages in the tidyverse. Together they can build a complex data science pipeline.  

Note that once we have loaded the tidyverse, we can call any of the functions from the individual packages. 
